---
title: "Data Visualisation Project - Study of wildfires and fire servise calls in Greece for summer 2021"
author: "Papadimitriou E. F."
date: "2023-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Data Visualization project - Study of forest fires in Greece and the human factor

We will try to combine satellite data about forest fires in Greece in the summer of 2021 with fire service calls data from the same period.

```{r, include=FALSE}
suppressWarnings({
library(dplyr)
library(readxl)
library(openxlsx)
library(janitor)
library(lubridate)
library(ggplot2)
library(scales)
library(ggmap)
library(maps)
library(patchwork)
library(tidyverse)
library(terra)
library(geodata)
library(maptiles)
library(raster)
library(sp)
library(sf)
library(leaflet)
library(leaflet.extras2)
library(shiny)
library(devtools)
library(osmdata)
library(RColorBrewer)
})

```


We are going to use VIIRS data for Greece in csv form, directly downloaded from: https://firms.modaps.eosdis.nasa.gov/country/

https://firms.modaps.eosdis.nasa.gov/data/country/viirs-snpp/2021/viirs-snpp_2021_Greece.csv

```{r}
#To get coordinates from satellite VIIRS csv
suppressWarnings({
viirs = read.csv("https://firms.modaps.eosdis.nasa.gov/data/country/viirs-snpp/2021/viirs-snpp_2021_Greece.csv", as.is=T)

})

viirs <- viirs |>
    mutate(a_month = format(as.Date(acq_date), "%m"),
         a_week = isoweek(as.Date(acq_date)),
         pixel_area_ha = 375*375/10000)


mymonths <- c("1.Jan","2.Feb","3.Mar",
              "4.Apr","5.May","6.Jun",
              "7.Jul","8.Aug","9.Sep",
              "10.Oct","11.Nov","12.Dec")
#add abbreviated month name
viirs$Month <- mymonths[ as.double(viirs$a_month) ]



for (i in 1:14) viirs$a_week[i] = 1
#Max and min coordinates for Greece:

upperLeft = c(41.91452561696354, 18.573755594482613)

lowerRight = c(35.06779648445501, 28.350020019193696)


#For prefectures we use a shapefile downloaded from:
#https://geodata.gov.gr/en/dataset/oria-nomon-okkhe

#Quick peek
head(viirs)

```



We will also add Forest Fire incident reports for 2021 by the Greek Fire Brigade from
https://www.fireservice.gr/el/synola-dedomenon

https://www.fireservice.gr/documents/20184/412680/Dasikes_Pyrkagies_2021.xls/d9b93e44-60b4-4a32-960b-84bbcdb1efe5
```{r}
#To get coordinates from fire department report log in xls
suppressWarnings({
#fdreports = read_xls("WildfiresGreece/Dasikes_Pyrkagies_2021.xls")

#fdreports = read.xlsx("https://www.fireservice.gr/documents/20184/412680/Dasikes_Pyrkagies#_2021.xls/d9b93e44-60b4-4a32-960b-84bbcdb1efe5", sheet = 1)

fdreports = read.xlsx("https://www.fireservice.gr/documents/20184/412680/Dasikes_Pyrkagies_2021.xlsx/0b3e94c3-d850-4604-ba63-21751434a99f", sheet = 1)

#Tidying colnames
fdreports_df <- as_data_frame(fdreports)
colnames(fdreports_df) <- fdreports_df[1,]
fdreports_df <- fdreports_df[-1,]
colnames(fdreports_df)[3] = "XENGAGE"
colnames(fdreports_df)[4] = "YENGAGE"
colnames(fdreports_df)[7] = "StartDate"
colnames(fdreports_df)[15] = "Woods"
colnames(fdreports_df)[16] = "WoodedArea"
colnames(fdreports_df)[17] = "Parks"
colnames(fdreports_df)[18] = "Grasslands"
colnames(fdreports_df)[19] = "Swamps"
colnames(fdreports_df)[20] = "Crops"
colnames(fdreports_df)[21] = "CropsRes"
colnames(fdreports_df)[22] = "Wastelands"
colnames(fdreports_df)[32] = "Helicopters"
colnames(fdreports_df)[33] = "CL415"
colnames(fdreports_df)[34] = "CL215"
colnames(fdreports_df)[35] = "PZL"
colnames(fdreports_df)[36] = "GRU."
colnames(fdreports_df)[37] = "HiredHeli"
colnames(fdreports_df)[38] = "HiredAircraft"

#Remove points outside Greece
fdreports_df <- filter(fdreports_df, XENGAGE > 18, YENGAGE > 34 )

#Tidying the excel dates
fdreports_df$StartDate <- as.numeric(as.character(fdreports_df$StartDate))
fdreports_df$StartDate <- as.Date(fdreports_df$StartDate, origin = "1899-12-30")
fdreports_df$'Ημερ/νία Κατασβεσης' <- as.numeric(as.character(fdreports_df$'Ημερ/νία Κατασβεσης'))
fdreports_df$'Ημερ/νία Κατασβεσης' <- as.Date(fdreports_df$'Ημερ/νία Κατασβεσης', origin = "1899-12-30")

#Clean lines with no coordinates for map creation
fdreports_clean <- fdreports_df |>
  drop_na("XENGAGE", "YENGAGE")
})

fdreports_clean$XENGAGE <- as.numeric(fdreports_clean$XENGAGE)
fdreports_clean$YENGAGE <- as.numeric(fdreports_clean$YENGAGE)


fdreports_clean <-fdreports_clean |>
  mutate(a_month = format(fdreports_clean$StartDate, "%m"),
         a_week = isoweek(as.Date(StartDate)),
         HiredHeli = coalesce(as.double(HiredHeli), 0),
         HiredAircraft = coalesce(as.double(HiredAircraft), 0),
         Helicopters = coalesce(as.double(Helicopters), 0),
         CL415 = coalesce(as.double(CL415), 0),
         CL215 = coalesce(as.double(CL215), 0),
         PZL = coalesce(as.double(PZL), 0),
         GRU. = coalesce(as.double(GRU.), 0),
         Woods = coalesce(as.double(Woods), 0),
         WoodedArea = coalesce(as.double(WoodedArea), 0),
         Parks = coalesce(as.double(Parks), 0),
         Grasslands = coalesce(as.double(Grasslands), 0),
         Swamps = coalesce(as.double(Swamps), 0),
         Crops = coalesce(as.double(Crops), 0),
         CropsRes = coalesce(as.double(CropsRes), 0),
         Wastelands = coalesce(as.double(Wastelands), 0)
         )

fdreports_clean <-fdreports_clean |>
  mutate(Woods = Woods*0.1,
         WoodedArea = WoodedArea*0.1,
         Parks = Parks*0.1,
         Grasslands = Grasslands*0.1,
         Swamps = Swamps*0.1,
         Crops = Crops*0.1,
         CropsRes = CropsRes*0.1,
         Wastelands = Wastelands*0.1
         )


#add abbreviated month name
fdreports_clean$Month <- mymonths[ as.double(fdreports_clean$a_month) ]
#points = st_as_sf(fdreports_clean, coords = c("XENGAGE", "YENGAGE"), crs = 4326)

#plot(st_geometry(points), pch=16, col="green")
#})

fdreports_clean$CL415[6506] = 6
fdreports_clean$CL215[6506] = 9
fdreports_clean$HiredHeli[7147] = 28
fdreports_clean$HiredAircraft[7147] = 9


fdreports_clean <-fdreports_clean |>
  mutate(AFairframes = Helicopters + CL415 + CL215 + PZL + GRU.,
         TotalAirframes = Helicopters + CL415 + CL215 + PZL + GRU. + HiredHeli + HiredAircraft,
         TotalIncLosses = Woods + WoodedArea + Parks + Grasslands + Swamps + Crops + CropsRes + Wastelands)

head(fdreports_clean |>
  dplyr::select(1:4, 7:10, 15:16, 32:34, 38:44))

# icon to blue square
square_blue <-
  makeIcon(iconUrl = "https://www.freeiconspng.com/uploads/blue-square-image-3.png",
           iconWidth = 18,
           iconHeight = 18)

#Color palette
#pal <- colorNumeric(palette = "YlOrRd", domain=ourShape@data$n)

# create leaflet map with active fire location points
fdreports2021_map <- leaflet(fdreports_clean) |>
  addTiles() |>
  addMarkers(lng = ~XENGAGE,
             lat = ~YENGAGE,
             icon = ~square_blue)

#fdreports2021_map

```

### Distribution of burned area in 2021

We plot the burned area distribution in the year using the burned area index by the fire service reports. There is burned area by the type of terrain, so we also create a column with the total area lost for each incident. We have tried to bin by week, so there might be multiple incidents stacked per week.  
```{r}
burnedarea <- fdreports_clean |>
  dplyr::select(3:4, 7, 15:22, TotalIncLosses, a_month, a_week, Month)

#For the plot based only on fd reports
suppressWarnings({
#pA <- burnedarea |>
#  ggplot()+
#  aes(x = a_week, y = as.double(TotalIncLosses), color = a_month)+
#  geom_col() +
#  labs(title = "Forest area lost per week of the year 2021 (by fire brigade reports)",
#       x = NULL,
#       y = "Area in  Ha",
#       color = NULL)+
#  theme(legend.position = "none") +
  #scale_y_continuous(labels = scales::comma, limits=c(0.1, 60000))
  #scale_y_continuous(trans = 'log10')
#  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
#              labels = trans_format("log10", math_format(10^.x)))

viirs |>
  ggplot()+
  aes(x = a_week, y=pixel_area_ha, fill = a_month)+
  geom_col() +
  labs(title = "Forest area lost per week of the year 2021 (by satellite)",
       x = NULL,
       y = "Area in Ha",
       color = NULL)+
  theme(legend.position = "none")+
  scale_y_continuous(labels = scales::comma, limits=c(0.1, 80000))
  #scale_y_continuous(trans = 'log10')
  #scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              #labels = trans_format("log10", math_format(10^.x)))


})
```



Another interesting idea is to check the type of land destroyed by incident
```{r}
burnedareaLong <- pivot_longer(burnedarea, cols = 4:11,
                              names_to = "LandType",
                              values_to = "AreaLost")

#Filter for summer
burnedareaLong <- burnedareaLong |>
  dplyr::filter(StartDate >= "2021-06-01", StartDate <= "2021-08-31")

burnedareaLong |>
  ggplot()+
  aes(x = StartDate, y = as.double(AreaLost), fill = LandType)+
  geom_col() +
  labs(title = "Daily area lost for the summer of 2021, by type of land (by fire brigade reports)",
       x = NULL,
       y = "Area in Ha",
       color = NULL)+
  #theme(legend.position = "plot") +
  scale_y_continuous(labels = scales::comma, limits=c(0.1, 60000))
  #scale_y_continuous(trans = 'log10')
  #scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              #labels = trans_format("log10", math_format(10^.x)))


```


### Use of airframes compared to the size of the fire

The next part of the analysis aims to deduce the possible tactics used by the Hellenic Fire Service. We expect the number and type of airframes used to be proportionate to the size of the forest fire. This we will check by plotting the area destroyed as a function of the total airframes used
```{r}
suppressWarnings({

p1 <-  ggplot(fdreports_clean)+
  aes(y = as.double(Woods), x = as.double(AFairframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Forest area in Ha",
       x = "Airframes used (HAF only)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  #scale_y_continuous(trans = 'log10') +
  xlim(1, 300)
  

p2 <-  ggplot(fdreports_clean)+
  aes(y = as.double(Woods), x = as.double(TotalAirframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Forest area in Ha",
       x = "Airframes used (total)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  #scale_y_continuous(trans = 'log10') +
  xlim(1, 300)

p3 <-  ggplot(fdreports_clean)+
  aes(y = as.double(TotalIncLosses), x = as.double(TotalAirframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Total area in Ha",
       x = "Airframes used (total)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  #scale_y_continuous(trans = 'log10') +
  xlim(1, 300)

p1 | p2 | p3
})
```

We see how data can sometimes lie. By plotting the "Wood" terrain compared to Air Force aircraft used we get a picture that does not seem linear at all. By changing that to include all used aircraft (Air Force and rented) as well as the entire area lost by any given forest fire (independent of terrain), we get something far more linear as expected.

### Types of aircraft used compared to the size of the fire

Continuing our research on tactics we also check the use of each separate airframe type compared to the size of the fire. While not explicitly described, we will put all available airframes in 5 categories, Helicopter,Light Aircraft, Large Aircraft or Other (unspecified types). We also add an extra layer, whether an aircraft is hired or not.
```{r}
#New table for airframes only
Airframes_df <- fdreports_clean |>
  dplyr::select(3:4, 7, 'TotalIncLosses', 32:42)

AirframesLong <- pivot_longer(Airframes_df, cols = 5:11,
                              names_to = "Type",
                              values_to = "Number")
AirframesLong <- AirframesLong |>
  mutate(Hired = ifelse(grepl("Hired", Type), "Hired", "HAF"),
         Category = ifelse(grepl("Heli", Type), "Helicopter", 
                           ifelse(grepl("CL", Type), "LargeAircraft", 
                                  ifelse(grepl("PZL", Type), "LightAircraft", 
                                         ifelse(grepl("HiredAir", Type), "LargeAircraft", "Other")))))

HAFonly <- fdreports_clean |>
  dplyr::select(3:4, 7, 'TotalIncLosses', 32:36, 40:42)

HAF_Long <- pivot_longer(Airframes_df, cols = 5:9,
                              names_to = "Type",
                              values_to = "Number")

HAF_Long <- HAF_Long |>
  mutate(Category = ifelse(grepl("Helicopters", Type), "Helicopter", 
                           ifelse(grepl("CL", Type), "LargeAircraft", 
                                  ifelse(grepl("PZL", Type), "LightAircraft", "Unknown type"))))


suppressWarnings({

ggplot(AirframesLong)+
  aes(y = TotalIncLosses, x = Number, color = Category, shape = Hired)+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Total area Ha",
       x = "Airframes used (Total)")+
  scale_y_continuous(labels = scales::comma)+
  #scale_y_continuous(trans = 'log10') +
  facet_wrap(~Category, scales = "free_y", ncol = 1)

})

```

From the above graphs, there is no clear indication of a unified strategy. What seems to be happening is that whatever system is close-by and available, is sent. While simplistic, it does make sense to simply send 1 or 2 aircraft of any type to start attacking the fire immediately, since you do not know how big it can get or how many others might appear, where more aircraft will be needed.

### Maps for the summer months of 2021

To start, we check the helicopter use for the whole country for the week of August.
```{r}
suppressMessages({
#Attempt to map from shapefile
#Idea from: https://gis.stackexchange.com/questions/460615/using-a-downloaded-shapefile-in-# ggplot2-and-r-readogr-error
#gr_shp <- st_read(file.path("C:/Users/manol/Documents/UniLu-MADS/Semester #1/DataVisualization/FinalProject/Solution/DataVisProject_Wildfires/WildfiresGreece/nomoiok#xe/nomoi_okxe", "nomoi_okxe.shp"))

# Basic plot of this shape file:
#par(mar=c(0,0,0,0))
#plot(st_geometry(gr_shp), col="darkgrey", bg="skyblue", lwd=0.25, border=0 )

#A better map
#Box for Greece
gr_borders = c(bottom  = 34.06779648445501, 
                 top     = 41.91452561696354,
                 left    = 18.573755594482613,
                 right   = 28.350020019193696)

#Create a stadia terrain (osm) map
pacman::p_load(ggmap, osmdata)
register_stadiamaps("98c88299-8c42-49e6-924d-98452750666e", write = FALSE)
gr_map <- get_stadiamap(gr_borders, zoom = 7, maptype = "stamen_terrain")

#Create a helicopter use table
AirframesLongH <- AirframesLong |>
  dplyr::filter(Category == "Helicopter")

AirframesLongH <- AirframesLongH |>
  dplyr::filter(Number != 0)

AirframesLongH <- AirframesLongH |>
  mutate(HelicopterNumber = Number)

AirframesLongH <- AirframesLongH |>
  dplyr::filter(a_week >= "31", a_week <= "34")

#Finally, a map of helicopter use density for the entire year
ggmap(gr_map) +
    geom_point(data = AirframesLongH, mapping = aes(x = XENGAGE, y = YENGAGE, 
                                        size = HelicopterNumber, alpha = 0.5)) +
    labs(title = "Helicopter use density in firefighting in Greece",
         subtitle = "Weeks of August, 2021",
         x=NULL,
         y=NULL) +
    theme(legend.position = "none",
          axis.text.x=element_blank(),
          axis.text.y=element_blank()) +
    facet_wrap(~a_week, ncol = 4)
})
```
```{r}
#Working on shapefile
gr_shp <- st_read(file.path("C:/Users/manol/Documents/UniLu-MADS/Semester 1/DataVisualization/FinalProject/Solution/DataVisProject_Wildfires/WildfiresGreece/oria_pyrosvestikwn_yphresiwn", "oria_pyrosvestikwn_yphresiwn.shp"))

#Basic plot of this shape file:
par(mar=c(0,0,0,0))
plot(st_geometry(gr_shp), col="darkgrey", bg="skyblue", lwd=0.25, border=0 )

gr_shp2 <- sf::st_read("C:/Users/manol/Documents/UniLu-MADS/Semester 1/DataVisualization/FinalProject/Solution/DataVisProject_Wildfires/WildfiresGreece/oria_pyrosvestikwn_yphresiwn/oria_pyrosvestikwn_yphresiwn.shp")

ggplot(gr_shp2) + 
  geom_sf(aes(fill = Area)) +
  theme_bw() -> gr_fd_map


```


Then we compare the reported burned areas by the Fire Service investigations to the live active fire spots by VIIRS instruments.
```{r}
suppressWarnings({

#Box for Evoia
ev_borders = c(bottom  = 37.8, 
                 top     = 39.2,
                 left    = 23,
                 right   = 25)

#Create a stadia terrain (osm) map
pacman::p_load(ggmap, osmdata)
register_stadiamaps("98c88299-8c42-49e6-924d-98452750666e", write = FALSE)
ev_map <- get_stadiamap(ev_borders, zoom = 9, maptype = "stamen_terrain")

#Create a burned area table for the 3 summer months
burnedarea_summer <- burnedarea |>
  dplyr::filter(StartDate >= "2021-06-01", StartDate <= "2021-08-31")

#Finally, a map of burned area for Euvoia for the summer
p4 <- ggmap(ev_map) +
    geom_point(data = burnedarea_summer, mapping = aes(x = XENGAGE, y = YENGAGE, 
                                        size = TotalIncLosses, color = "red", alpha = 0.5)) +
    labs(title = "Burned area reports by Fire Service in the island of Euvoia",
         subtitle = "June-August 2021",
         x=NULL,
         y=NULL) +
    theme(axis.text.x=element_blank(),
          axis.text.y=element_blank()) +
    facet_wrap(~Month, ncol = 3)



#Create an active fire tble for the 3 summer months

activefires <- viirs |>
  dplyr::filter(acq_date >= "2021-06-01", acq_date <= "2021-08-31")

#Finally, a map of active fires for Euvoia for the summer
p5 <- ggmap(ev_map) +
    geom_point(data = activefires, mapping = aes(x = longitude, y = latitude, 
                                        col = confidence)) +
    labs(title = "Active fire instances by VIIRS in the island of Euvoia",
         subtitle = "June-August 2021",
         x=NULL,
         y=NULL) +
    theme(axis.text.x=element_blank(),
          axis.text.y=element_blank()) +
    facet_wrap(~Month, ncol = 3) +
    scale_color_brewer(palette = "YlOrRd")

p4/p5
})
```


