---
title: "Data Visualisation Project - Study of wildfires and fire servise calls in Greece for summer 2021"
author: "Papadimitriou E. F."
date: "2023-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Data Visualization project - Study of forest fires in Greece and the human factor

We will try to combine satellite data about forest fires in Greece in the summer of 2021 with fire service calls data from the same period.

```{r, include=FALSE}
suppressWarnings({
library(dplyr)
library(readxl)
library(janitor)
library(lubridate)
library(ggplot2)
library(ggmap)
library(maps)
library(patchwork)
library(tidyverse)
library(terra)
library(geodata)
library(maptiles)
library(raster)
library(sp)
library(sf)
library(leaflet)
library(leaflet.extras2)
library(shiny)
library(devtools)
library(osmdata)
library(RColorBrewer)
})

```


We are going to use VIIRS data for Greece in csv form, directly downloaded from: https://firms.modaps.eosdis.nasa.gov/country/

```{r}
#To get coordinates from satellite VIIRS csv
suppressWarnings({
viirs = read.csv("C:/Users/manol/Documents/UniLu-MADS/Semester 1/DataVisualization/FinalProject/Solution/DataVisProject_Wildfires/WildfiresGreece/viirs-snpp_2021_Greece.csv", as.is=T)

})

viirs <- viirs |>
    mutate(a_month = format(as.Date(acq_date), "%m"),
         a_week = isoweek(as.Date(acq_date)))


for (i in 1:14) viirs$a_week[i] = 1
#Max and min coordinates for Greece:

upperLeft = c(41.91452561696354, 18.573755594482613)

lowerRight = c(35.06779648445501, 28.350020019193696)


#For prefectures we use a shapefile downloaded from:
#https://geodata.gov.gr/en/dataset/oria-nomon-okkhe

#Quick peek
head(viirs)

```



We will also add Forest Fire incident reports for 2021 by the Greek Fire Brigade from
https://www.fireservice.gr/el/synola-dedomenon

```{r}
#To get coordinates from fire department report log in xls
suppressWarnings({
fdreports = read_xls("WildfiresGreece/Dasikes_Pyrkagies_2021.xls", .name_repair = "universal")

#Tidying colnames
fdreports_df <- as_data_frame(fdreports)
colnames(fdreports_df) <- fdreports_df[1,]
fdreports_df <- fdreports_df[-1,]
colnames(fdreports_df)[3] = "XENGAGE"
colnames(fdreports_df)[4] = "YENGAGE"
colnames(fdreports_df)[7] = "StartDate"
colnames(fdreports_df)[15] = "Woods"
colnames(fdreports_df)[16] = "WoodedArea"
colnames(fdreports_df)[17] = "Parks"
colnames(fdreports_df)[18] = "Grasslands"
colnames(fdreports_df)[19] = "Swamps"
colnames(fdreports_df)[20] = "Crops"
colnames(fdreports_df)[21] = "CropsRes"
colnames(fdreports_df)[22] = "Wastelands"
colnames(fdreports_df)[32] = "Helicopters"
colnames(fdreports_df)[33] = "CL415"
colnames(fdreports_df)[34] = "CL215"
colnames(fdreports_df)[35] = "PZL"
colnames(fdreports_df)[36] = "GRU."
colnames(fdreports_df)[37] = "HiredHeli"
colnames(fdreports_df)[38] = "HiredAircraft"

#Remove points outside Greece
fdreports_df <- filter(fdreports_df, XENGAGE > 18, YENGAGE > 34 )

#Tidying the excel dates
fdreports_df$StartDate <- as.numeric(as.character(fdreports_df$StartDate))
fdreports_df$StartDate <- as.Date(fdreports_df$StartDate, origin = "1899-12-30")
fdreports_df$'Ημερ/νία Κατασβεσης' <- as.numeric(as.character(fdreports_df$'Ημερ/νία Κατασβεσης'))
fdreports_df$'Ημερ/νία Κατασβεσης' <- as.Date(fdreports_df$'Ημερ/νία Κατασβεσης', origin = "1899-12-30")

#Clean lines with no coordinates for map creation
fdreports_clean <- fdreports_df |>
  drop_na("XENGAGE", "YENGAGE")
})

fdreports_clean$XENGAGE <- as.numeric(fdreports_clean$XENGAGE)
fdreports_clean$YENGAGE <- as.numeric(fdreports_clean$YENGAGE)


fdreports_clean <-fdreports_clean |>
  mutate(a_month = format(fdreports_clean$StartDate, "%m"),
         a_week = isoweek(as.Date(StartDate)),
         HiredHeli = coalesce(as.double(HiredHeli), 0),
         HiredAircraft = coalesce(as.double(HiredAircraft), 0),
         Helicopters = coalesce(as.double(Helicopters), 0),
         CL415 = coalesce(as.double(CL415), 0),
         CL215 = coalesce(as.double(CL215), 0),
         PZL = coalesce(as.double(PZL), 0),
         GRU. = coalesce(as.double(GRU.), 0),
         Woods = coalesce(as.double(Woods), 0),
         WoodedArea = coalesce(as.double(WoodedArea), 0),
         Parks = coalesce(as.double(Parks), 0),
         Grasslands = coalesce(as.double(Grasslands), 0),
         Swamps = coalesce(as.double(Swamps), 0),
         Crops = coalesce(as.double(Crops), 0),
         CropsRes = coalesce(as.double(CropsRes), 0),
         Wastelands = coalesce(as.double(Wastelands), 0)
         )
#points = st_as_sf(fdreports_clean, coords = c("XENGAGE", "YENGAGE"), crs = 4326)

#plot(st_geometry(points), pch=16, col="green")
#})

fdreports_clean$CL415[6506] = 6
fdreports_clean$CL215[6506] = 9
fdreports_clean$HiredHeli[7147] = 28
fdreports_clean$HiredAircraft[7147] = 9


fdreports_clean <-fdreports_clean |>
  mutate(AFairframes = Helicopters + CL415 + CL215 + PZL + GRU.,
         TotalAirframes = Helicopters + CL415 + CL215 + PZL + GRU. + HiredHeli + HiredAircraft,
         TotalIncLosses = Woods + WoodedArea + Parks + Grasslands + Swamps + Crops + CropsRes + Wastelands)

head(fdreports_clean |>
  dplyr::select(1:4, 7:10, 15:16, 32:34, 38:43))

# icon to blue square
square_blue <-
  makeIcon(iconUrl = "https://www.freeiconspng.com/uploads/blue-square-image-3.png",
           iconWidth = 18,
           iconHeight = 18)

#Color palette
#pal <- colorNumeric(palette = "YlOrRd", domain=ourShape@data$n)

# create leaflet map with active fire location points
fdreports2021_map <- leaflet(fdreports_clean) |>
  addTiles() |>
  addMarkers(lng = ~XENGAGE,
             lat = ~YENGAGE,
             icon = ~square_blue)

#fdreports2021_map

```

### Distribution of burned area in 2021

We plot the burned area distribution in the year using the burned area index by the fire service reports. There is burned area by the type of terrain, so we also create a column with the total area lost for each incident. We have tried to bin by week, so there might be multiple incidents stacked per week.  
```{r}
burnedarea <- fdreports_clean |>
  dplyr::select(7, Woods, TotalIncLosses, a_month, a_week)

#For the plot based only on fd reports
#suppressWarnings({
burnedarea |>
  ggplot()+
  aes(x = a_week, y = as.double(TotalIncLosses), color = a_month)+
  geom_col() +
  labs(title = "Forest area lost per week of the year 2021 (by fire brigade reports)",
       x = NULL,
       y = "Area in 1,000 sq. meters (0.1 Ha)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma, limits=c(0.1, 600000))
#})
```


### Use of airframes compared to the size of the fire

The next part of the analysis aims to deduce the possible tactics used by the Hellenic Fire Service. We expect the number and type of airframes used to be proportionate to the size of the forest fire. This we will check by plotting the area destroyed as a function of the total airframes used
```{r}
suppressWarnings({

p1 <-  ggplot(fdreports_clean)+
  aes(y = as.double(Woods), x = as.double(AFairframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Forest area in 1,000 sq. meters (0.1 Ha)",
       x = "Airframes used (HAF only)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  xlim(1, 300)
  

p2 <-  ggplot(fdreports_clean)+
  aes(y = as.double(Woods), x = as.double(TotalAirframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Forest area in 1,000 sq. meters (0.1 Ha)",
       x = "Airframes used (total)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  xlim(1, 300)

p3 <-  ggplot(fdreports_clean)+
  aes(y = as.double(TotalIncLosses), x = as.double(TotalAirframes))+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Total area in 1,000 sq. meters (0.1 Ha)",
       x = "Airframes used (total)",
       color = NULL)+
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma) +
  xlim(1, 300)

p1 | p2 | p3
})
```

We see how data can sometimes lie. By plotting the "Wood" terrain compared to Air Force aircraft used we get a picture that does not seem linear at all. By changing that to include all used aircraft (Air Force and rented) as well as the entire area lost by any given forest fire (independent of terrain), we get something far more linear as expected.

##Types of aircraft used compared to the size of the fire

Continuing our research on tactics we also check the use of each separate airframe type compared to the size of the fire. While not explicitly described, we will put all available airframes in 5 categories, Helicopter,Light Aircraft, Large Aircraft or Other (unspecified types). We also add an extra layer, whether an aircraft is hired or not.
```{r}
#New table for airframes only
Airframes_df <- fdreports_clean |>
  dplyr::select(3:4, 7, 'TotalIncLosses', 32:41)

AirframesLong <- pivot_longer(Airframes_df, cols = 5:11,
                              names_to = "Type",
                              values_to = "Number")
AirframesLong <- AirframesLong |>
  mutate(Hired = ifelse(grepl("Hired", Type), "Hired", "HAF"),
         Category = ifelse(grepl("Heli", Type), "Helicopter", 
                           ifelse(grepl("CL", Type), "LargeAircraft", 
                                  ifelse(grepl("PZL", Type), "LightAircraft", 
                                         ifelse(grepl("HiredAir", Type), "LargeAircraft", "Other")))))

HAFonly <- fdreports_clean |>
  dplyr::select(3:4, 7, 'TotalIncLosses', 32:36, 40:41)

HAF_Long <- pivot_longer(Airframes_df, cols = 5:9,
                              names_to = "Type",
                              values_to = "Number")

HAF_Long <- HAF_Long |>
  mutate(Category = ifelse(grepl("Helicopters", Type), "Helicopter", 
                           ifelse(grepl("CL", Type), "LargeAircraft", 
                                  ifelse(grepl("PZL", Type), "LightAircraft", "Other"))))


suppressWarnings({

ggplot(AirframesLong)+
  aes(y = TotalIncLosses, x = Number, color = Category, shape = Hired)+
  geom_point() +
  geom_smooth(method = "loess", span = 5, formula = 'y ~ x') +
  labs(title = NULL,
       y = "Total area in 1,000 sq. meters (0.1 Ha)",
       x = "Airframes used (Total)")+
  scale_y_continuous(labels = scales::comma)+
  facet_wrap(~Category, scales = "free_y", ncol = 1)

})

```

From the above graphs, there is no clear indication of a unified strategy. What seems to be happening is that whatever system is close-by and available, is sent. While simplistic, it does make sense to simply send 1 or 2 aircraft of any type to start attacking the fire immediately, since you do not know how big it can get or how many others might appear, where more aircraft will be needed.

###Several maps for the summer months

Non-interactive helicopter use map for Summer 2021 (Working)
```{r}
#Attempt to map from shapefile
#Idea from: https://gis.stackexchange.com/questions/460615/using-a-downloaded-shapefile-in-# ggplot2-and-r-readogr-error
gr_shp <- st_read(file.path("C:/Users/manol/Documents/UniLu-MADS/Semester 1/DataVisualization/FinalProject/Solution/DataVisProject_Wildfires/WildfiresGreece/nomoiokxe/nomoi_okxe", "nomoi_okxe.shp"))

# Basic plot of this shape file:
#par(mar=c(0,0,0,0))
#plot(st_geometry(gr_shp), col="darkgrey", bg="skyblue", lwd=0.25, border=0 )

#A better map
#Box for Greece
gr_borders = c(bottom  = 34.06779648445501, 
                 top     = 41.91452561696354,
                 left    = 18.573755594482613,
                 right   = 28.350020019193696)

#Create a stadia terrain (osm) map
pacman::p_load(ggmap, osmdata)
register_stadiamaps("98c88299-8c42-49e6-924d-98452750666e", write = FALSE)
gr_map <- get_stadiamap(gr_borders, zoom = 7, maptype = "stamen_terrain")

#Create a helicopter use table
AirframesLongH <- AirframesLong |>
  dplyr::filter(Category == "Helicopter")

AirframesLongH <- AirframesLongH |>
  dplyr::filter(Number != 0)

AirframesLongH <- AirframesLongH |>
  mutate(HelicopterNumber = Number)

AirframesLongH <- AirframesLongH |>
  dplyr::filter(StartDate >= "2021-06-01", StartDate <= "2021-08-31")

#Finally, a map of helicopter use density for the entire year
ggmap(gr_map) +
    geom_point(data = AirframesLongH, mapping = aes(x = XENGAGE, y = YENGAGE, 
                                        size = HelicopterNumber, alpha = 0.5)) +
    labs(title = "Helicopter use density in firefighting in Greece",
         subtitle = "Summer 2021",
         x=NULL,
         y=NULL) +
    theme(legend.position = "none") +
    facet_wrap(~a_month, ncol = 3)
```


Non-interactive active fire map for mega fire in the island of Euvoia, August 2021
```{r}
#Attempt to map from shapefile
#Idea from: https://gis.stackexchange.com/questions/460615/using-a-downloaded-shapefile-in-# ggplot2-and-r-readogr-error

#Box for Evoia
ev_borders = c(bottom  = 37.8, 
                 top     = 39.2,
                 left    = 23,
                 right   = 25)

#Create a stadia terrain (osm) map
pacman::p_load(ggmap, osmdata)
register_stadiamaps("98c88299-8c42-49e6-924d-98452750666e", write = FALSE)
ev_map <- get_stadiamap(ev_borders, zoom = 9, maptype = "stamen_terrain")

#Create a helicopter use table

activefires <- viirs |>
  dplyr::filter(acq_date >= "2021-06-01", acq_date <= "2021-08-31")

#Finally, a map of helicopter use density for the entire year
ggmap(ev_map) +
    geom_point(data = activefires, mapping = aes(x = longitude, y = latitude, 
                                        col = confidence)) +
    labs(title = "Active fire instances by VIIRS in the island of Euvoia",
         subtitle = "Summer 2021",
         x=NULL,
         y=NULL) +
    theme(legend.position = "none") +
    facet_wrap(~a_month, ncol = 3) +
    scale_color_brewer(palette = "YlOrRd")
```

Mapping out different airframe use for the 3 summer months (Shiny App)

First, for helicopters
```{r}
#We will check July for starters
#AirframesLongJuly <- AirframesLong |>
#  dplyr::filter(a_month == "07")

#We will only check helicopters for simplicity
AirframesLongH <- AirframesLong |>
  dplyr::filter(Category == "Helicopter")

AirframesLongH <- AirframesLongH |>
  dplyr::filter(Number != 0)

#Icon for helicopters
helicopter_icon <-
  makeIcon(iconUrl = "https://www.freeiconspng.com/uploads/helicopter-icon-17.png",
           iconWidth = 18,
           iconHeight = 18)

#Map of uses of the specific airframe in July

helicopter_July_map <- leaflet(width = "100%") %>%
  setView(lng = 24.16, lat = 38.51, zoom = 6) %>%
  addProviderTiles("Esri.WorldStreetMap")

# specify shiny user interface and time slider


#https://blogs.unimelb.edu.au/researcher-library/2020/06/19/making-a-covid-19-map-in-r-usin#g-shiny-and-leaflet/

#BETTER
#https://stackoverflow.com/questions/69276453/how-do-i-create-a-shiny-slider-app-to-view-my-data-over-time-on-the-map-that-i-h
ui <- shinyUI(pageWithSidebar(
  headerPanel("Map of Helicopter use for Summer 2021"),
  sidebarPanel(
    dateRangeInput(
      "range",
      "Date range",
      start  = "2021-06-01",
      end    = "2021-08-31"
    )),
  mainPanel(
    leafletOutput("map")
  )
))

# shiny server input/output
server <- shinyServer(function(input, output) {
  filtered_data_H <- reactive({
    AirframesLongH |> filter(AirframesLongH$StartDate >= input$range[1] & AirframesLongH$StartDate <= input$range[2])
  })

  output$map <- renderLeaflet({
    helicopter_July_map |>
      addMarkers(
        data = filtered_data_H(),
        lng = ~XENGAGE,
        lat = ~YENGAGE,
        icon = ~helicopter_icon
      ) 
      
  })
})

shinyApp(ui, server)

```



Then, for heavy aircraft:
```{r}
  #We will only check heavy aircraft for simplicity
AirframesLongAXL <- AirframesLong |>
  dplyr::filter(Category == "LargeAircraft")

AirframesLongAXL <- AirframesLongAXL |>
  dplyr::filter(Number != 0)

#Icon for helicopters
AXL_icon <-
  makeIcon(iconUrl = "https://www.freeiconspng.com/uploads/plane-icon--iconshow-14.png",
           iconWidth = 18,
           iconHeight = 18)

#Map of uses of the specific airframe in July

HeavyAircraft_July_map <- leaflet(width = "100%") %>%
  setView(lng = 24.16, lat = 38.51, zoom = 6) %>%
  addProviderTiles("Esri.WorldStreetMap")

# specify shiny user interface and time slider


#https://blogs.unimelb.edu.au/researcher-library/2020/06/19/making-a-covid-19-map-in-r-usin#g-shiny-and-leaflet/

#BETTER
#https://stackoverflow.com/questions/69276453/how-do-i-create-a-shiny-slider-app-to-view-my-data-over-time-on-the-map-that-i-h
ui2 <- shinyUI(pageWithSidebar(
  headerPanel("Map of Large Aircraft use for Summer 2021"),
  sidebarPanel(
    dateRangeInput(
      "range",
      "Date range",
      start  = "2021-06-01",
      end    = "2021-08-31"
    )),
  mainPanel(
    leafletOutput("map")
  )
))

# shiny server input/output
server2 <- shinyServer(function(input, output) {
  filtered_data_AXL <- reactive({
    AirframesLongAXL |> filter(AirframesLongAXL$StartDate >= input$range[1] & AirframesLongAXL$StartDate <= input$range[2])
  })

  output$map <- renderLeaflet({
    HeavyAircraft_July_map |>
      addMarkers(
        data = filtered_data_AXL(),
        lng = ~XENGAGE,
        lat = ~YENGAGE,
        icon = ~AXL_icon
      ) 
      
  })
})

shinyApp(ui2, server2)

```



For active fires by VIIRS satelite data
```{r}
#We only choose high confidence squares
activefires <- viirs |>
  dplyr::filter(confidence == "h")


square_red <-
  makeIcon(iconUrl = "https://www.freeiconspng.com/uploads/red-square-png-14.png",
           iconWidth = 18,
           iconHeight = 18)

livefires_map <- leaflet(width = "100%") %>%
  setView(lng = 24.16, lat = 38.51, zoom = 6) %>%
  addProviderTiles("Esri.WorldStreetMap")

# specify shiny user interface and time slider
ui3 <- shinyUI(pageWithSidebar(
  headerPanel("Map of Active fires by VIIRS sensors for Summer 2021"),
  sidebarPanel(
    dateRangeInput(
      "range",
      "Date range",
      start  = "2021-06-01",
      end    = "2021-08-31"
    )),
  mainPanel(
    leafletOutput("map")
  )
))

# shiny server input/output
server3 <- shinyServer(function(input, output) {
  filtered_data_active <- reactive({
    activefires |> filter(activefires$acq_date >= input$range[1] & activefires$acq_date <= input$range[2])
  })

  output$map <- renderLeaflet({
    livefires_map |>
      addMarkers(
        data = filtered_data_active(),
        lng = ~longitude,
        lat = ~latitude,
        icon = ~square_red
      ) 
      
  })
})

shinyApp(ui3, server3)
```

